\lecture{Lecture 14: Face Detection}

\qa{Why is the sliding window approach computationally daunting for face detection, and what two absolute design imperatives must the Viola-Jones architecture satisfy to overcome this?}
{
\begin{itemize}
    \item The sliding window must evaluate all spatial locations across multiple scales, resulting in millions of candidate patches per image.
    \item Since the vast majority of these patches are non-faces, the architecture must achieve two things to be viable: 1) \textbf{extremely fast processing} to rapidly discard non-face background patches, and 2) a \textbf{very low false positive rate} to prevent the sheer volume of background windows from overwhelming the final output.
\end{itemize}}

\qa{How are Haar features mathematically defined and evaluated, and what is the engineering trade-off regarding their structural coarseness?}
{
\begin{itemize}
    \item A Haar feature is fundamentally defined as the strictly coupled combination of a \textbf{function} and a \textbf{geometry}:
    \begin{itemize}
        \item \textbf{Function:} The mathematical operation $f(x)$, which is the scalar difference between the sum of pixel intensities in adjacent "black" and "white" rectangular regions:
        $$f(x) = \sum_i p_{black}(i) - \sum_i p_{white}(i)$$
        \item \textbf{Geometry:} The specific spatial layout (e.g., 2-rectangle, 3-rectangle), scale (width and height), and discrete position of those rectangles within the evaluation window.
    \end{itemize}
    \item \textbf{Trade-off:} Haar features are structurally coarse; they can only capture simple gradients like horizontal/vertical edges or bars. However, because their geometry is strictly rectangular, they can be computed in $O(1)$ time using an \textbf{integral image}. This massive computational efficiency perfectly compensates for their lack of fine structural detail.
\end{itemize}}

\qa{Given an intractable pool of \texorpdfstring{$\sim$}{\textasciitilde}160,000 possible Haar features in a \texorpdfstring{$24 \times 24$}{24x24} window, how does AdaBoost construct a strong classifier and force the algorithm to learn difficult edge cases?}
{
\begin{itemize}
    \item AdaBoost sequentially selects the single best "weak learner" at each iteration. A weak classifier evaluates a single Haar feature $f_j(x)$, optimized with a threshold $\theta_j$ and a parity $p_j$ (direction of the inequality):
    $$h_j(x) = \begin{cases} 1 & \text{if } p_j f_j(x) > p_j \theta_j \\ -1 & \text{otherwise} \end{cases}$$
    \item \textbf{Re-weighting mechanism:} After each round, training examples that were misclassified by the current weak learner have their weights exponentially increased, while correctly classified examples are down-weighted. This forces the subsequent weak classifiers in the next round to mathematically prioritize correcting previous mistakes. The final strong classifier is the sign of the weighted sum of all chosen weak learners: $h(x) = \text{sign}\left(\sum_{j=1}^{T} \alpha_j h_j(x)\right)$.
\end{itemize}}

\qa{What is the structural logic of the Viola-Jones cascade, and how does it enforce error asymmetry while drastically reducing computational cost?}
{
\begin{itemize}
    \item The cascade arranges weak classifiers into sequential, increasingly complex stages (e.g., Stage 1 uses 1 feature, Stage 2 uses 10 features). A sliding window must pass \textit{every} stage to be classified as a face; if it fails any stage, it is instantly rejected and computation halts.
    \item \textbf{Computational Efficiency (Early Rejection):} Since most sliding windows are purely background, the cascade rejects them extremely early. Rejecting a background patch in Stage 1 requires almost zero computation, saving the system from evaluating thousands of features on empty space. The complex later stages are strictly reserved for rare face-like regions.
    \item \textbf{Error Asymmetry:} A \textbf{False Positive} at an early stage is acceptable because stricter stages later on will filter it out. Conversely, a \textbf{False Negative} is fatal: if a true face is rejected early, it vanishes from the pipeline forever. Thus, early stages are aggressively tuned for near 100\% recall (catching all possible faces) at the heavy expense of precision.
\end{itemize}}

\qa{What visual cues does the Viola-Jones detector capture, and how do local Haar patterns relate to human facial components?}
{
\begin{itemize}
    \item Haar features capture broad, high-contrast intensity differences between adjacent regions rather than fine textures. For human faces, these differences naturally arise from the 3D structure of the face casting predictable shadows under normal lighting.
    \item The original paper notes that AdaBoost autonomously selected these specific geometric cues as the most universally distinctive. The first two features are:
    \begin{enumerate}
        \item A \textbf{vertical two-rectangle feature} measuring the difference between the \textbf{eye region} (which is darker due to \textbf{structural eye socket shadows}) and the \textbf{upper cheeks} (which protrude and catch light).
        \item A \textbf{horizontal three-rectangle feature} (side-by-side) contrasting the darker \textbf{eyes} on the exterior against the brighter, protruding \textbf{bridge of the nose} in the center.
    \end{enumerate}
    \item \textbf{Vulnerability \& Bias:} Since these features rely on structural 3D shadows rather than textures, they assume standard top-down lighting. If a dark-skinned person with light eyes under flat lighting breaks this intensity gradient, the first stage may output a negative response. Rejecting this at Stage 1 triggers an immediate, unrecoverable \textbf{False Negative} (causing the known racial and lighting biases in early systems).
\end{itemize}}